---
title: "Import data"
output: html_notebook
---

<!-- This file by Martin Monkman is licensed under a Creative Commons Attribution 4.0 International License. 
Some of it comes straight from Hadley Wickham & Garrett Grolemund's book _R for Data Science_, found online here: https://r4ds.had.co.nz/
It uses great swaths of content from the tidyverse reference material for {readr} and {readxl}, by RStudio. 
You can find that material here: 
https://readr.tidyverse.org/index.html 
https://readxl.tidyverse.org/index.html
-->


# Getting started

Run this code chunk:

```{r}

library(tidyverse)  # we will be using the functions of {readr}
library(readxl)

# utilities
library(here)
library(janitor)

```



# 1. Reading a CSV file

> Background reading: [11 Data import](https://r4ds.had.co.nz/data-import.html) in _R for Data Science_



CSV files are very commonly used for storing flat files. They don't have any formatting--just the number or text in the cell.

The package {readr} is designed to make importing these files simple.


This example uses data provided by Statistics Canada, with household income by household type from the 2016 Census. In this case, the file contains all of the census tracts in the Census Metropolitan Areas (CMAs) and Census Agglomerations (CAs). See [](..\scr\census_data_sources.Rmd) for more information. 

Run this chunk to create an object called `df_98_400_X2016100`, from a CSV file of the same name.

> Sidebar: the {here} package is designed to work with R projects. It defines the root of your project directory, and then the `here()` function uses a string that is not dependent on your operating system (slash vs. backslash to separate folders!) This helps with collaboration, and makes your code more reproducible.


```{r}
# example

df_98_400_X2016100 <- read_csv(here("data", "98-400-X2016100_ENG_CSV", "98-400-X2016100_English_CSV_data.csv"))

```

Oh no...there are parsing failures.

>1815 parsing failures.
 row                                                                                                 col expected actual                                                                                           file
1167 Dim: Household income statistics (3): Member ID: [1]: Total - Household income statistics (Note: 3) a double      x 'E:/github/SFU_CDSC103_2020/data/98-400-X2016100_ENG_CSV/98-400-X2016100_English_CSV_data.csv'
1167 Dim: Household income statistics (3): Member ID: [2]: Median total income of households ($)         a double      x 


This is something that you will sometimes see in Statistics Canada's files. Due to data quality and/or confidentiality, values are suppressed. In this case, the suppressed values are represented by an "x". This is problematic, because the column is of numeric values but the "x" is a character.

{readr} provides a solution to this ... there's an argument in the `read_csv()` function to define characters as "NA" (how R represents missing values).

## Your Turn 1.1

Insert an R chunk and rerun the example above, but with argument `na = "x" ` added.

```{r}
# solution

df_98_400_X2016100 <- read_csv(here("data", "98-400-X2016100_ENG_CSV", "98-400-X2016100_English_CSV_data.csv"),
                               na = "x")

```

Adding the `cols()` allows us to alter what {readr} has decided for us. For example, the variable `Member ID: Household type including census family structure (11)` is a categorical variable that describes the household type. {readr} has looked at it and decided that it's a number (a <dbl>), but we could set it to be an integer.

```{r}
# example
read_csv(here("data", "98-400-X2016100_ENG_CSV", "98-400-X2016100_English_CSV_data.csv"))



# example
read_csv(here("data", "98-400-X2016100_ENG_CSV", "98-400-X2016100_English_CSV_data.csv"),
                               col_types =
                                 cols(`Member ID: Household type including census family structure (11)` = col_integer()))


```





The {readr} package allows a lot of control over how the file is read. Of particular utility are 

* `na = ""` -- specify which values you want to be turned into `NA`

* `skip = 0` -- specify how many rows to skip 

* `n_max = Inf` -- the maximum number of records to read


## Your Turn 1.2

Read the first 5 rows of the "98-400-X2016100_English_CSV_data.csv" file.

```{r}
# solution
read_csv(here("data", "98-400-X2016100_ENG_CSV", "98-400-X2016100_English_CSV_data.csv"),
         n_max = 5)


```




STOP HERE ...


# 2. Reading an Excel file

If anything, Excel files are more common than CSV and other plain-text data files. They seem to multiply like coat hangers in the closet...

And as pointed out in the article by Karl Broman & Kara Woo, ["Data Organization in Spreadsheets"](https://www.tandfonline.com/doi/full/10.1080/00031305.2017.1375989), they can encourage some ways of storing information that makes it hard for us to analyze. Excel files can also contain a wide variety of data format types. {readxl} tries to figure out what's going on, but like {readr}, it allows you to override some of those automatic decisions.


```{r}
deaths <- read_excel("data/deaths.xlsx")
deaths
```

What do you notice about the "Date of birth" and "Date of death" columns?


## Your Turn 2.1

Read in the "deaths" file, but use the `range = ` option to specify when to start reading the file:


```{r}
# Example
deaths <- read_excel("data/deaths.xlsx", 
                     range = "A1:B3")

deaths

# solution
deaths <- read_excel(
    "data/deaths.xlsx",
    range = "A5:F15"
  )

deaths

```


Use the `skip =` and `n_max =` options to achieve the same thing:

```{r}

# solution
deaths <- read_excel(
    "data/deaths.xlsx",
    skip = 4,
    n_max = 10
  )

deaths


```


## Your Turn 2.2

Excel files often (almost always?) have multiple sheets in them. It's possible to specify which one you want to use with the `sheets = ""` option. You can also use the position number (if you happen to know it). 

Note that if you don't specify the sheet, {readxl} will default to the first one.

The `excel_sheets()` function will tell you the name of the sheets in an Excel file.

```{r}
excel_sheets("data/deaths.xlsx")

```

Now, read in the "other" sheet using the name of the sheet.


```{r}

# Solution
read_excel("data/deaths.xlsx", "other")


```

And finally, read in the "other" sheet, using the position.

```{r}
# Solution
read_excel("data/deaths.xlsx", 2)

```




## REFERENCE MATERIAL

You can find more details in _R for Data Science_:
https://r4ds.had.co.nz/data-import.html


{readr} 
https://readr.tidyverse.org/

{readxl}
https://readxl.tidyverse.org/index.html


