---
title: "Transform Data 1: manipulate cases"
output: html_notebook
---

<!-- This file by Charlotte Wickham (with some modifications by Martin Monkman) is licensed under a Creative Commons Attribution 4.0 International License, adapted from the orignal work at https://github.com/rstudio/master-the-tidyverse by RStudio. -->

```{r setup}
library(tidyverse)
library(gapminder)

```

This exercise introduces functions within the {dplyr} package. Note that the functions are (almost always) verbs, describing what the function does.

|function |action |
| :--- | ---: |
| `filter()`    | *keep rows*|
| `mutate()`    | *create a new variable*|
| `group_by()`  | *declare subsets in data*|
| `select()`    | *keep variables (or drop them -var)*|
| `rename()`    | *renaming variables*|
| `summarize()` | *summarize the data, by groups if they have been declared*|
| `tally()`     | *counting (by groups if group_by() applied)*|
| `distinct()`  | *returns only rows that are unique*| 
| `case_when()` | *is used for "recoding" variable, often used with mutate()*|



## gapminder

For this exercise, we will use data from the {gapminder} package. For each of 142 countries, the package provides values for life expectancy, GDP per capita, and population, every five years, from 1952 to 2007.

It's a slice of data found at [gapminder.org](gapminder.org), which was founded by Hans Rosling. If you haven't already seen his videos, I suggest:

* [200 Years That Changed The World](https://www.gapminder.org/videos/200-years-that-changed-the-world/)

* [TED talk "The Best Stats You've Ever Seen"](https://www.ted.com/talks/hans_rosling_the_best_stats_you_ve_ever_seen)

and his book, [_Factfulness: Ten Reasons Why We're Wrong About the World--and Why Things Are Better Than You Think_](https://www.gapminder.org/factfulness-book/).


First, we load the data. Since the package is already installed into our R environment, we can just do the following:

```{r}
gapminder
```

## Your Turn 1: filter() rows

See if you can use the logical operators to manipulate our code below to show:

The data for Canada

```{r}
filter(gapminder, country == "New Zealand")
```

All data for countries in Oceania

```{r}
filter(gapminder, country == "New Zealand")
```

Rows where the life expectancy is greater than 82
```{r}
filter(gapminder, country == "New Zealand")
```


## Your Turn 2: multiple criteria

Use Boolean operators to alter the code below to return only the rows that contain:

* United States before 1970

```{r}

gapminder %>%
  filter(country == "New Zealand", 
         year > 2000)

```

*  Countries where life expectancy in 2007 is below 50 or over 75

```{r}


```



* Create a list object and return records for any of "New Zealand", "Canada" or "United States"


```{r}

```

* Return records for all continents except Europe


```{r}

```

## Your Turn 3: sorting

Find the records with the smallest population.
```{r}

```

Find the records with the largest GDP per capita.
```{r}

```


## Your Turn 4: creating new variables

Calculate a new variable with total GDP

```{r, error = TRUE}



```

Create a summary table with the population and GDP by continent for the year 2007  

```{r}

```

## Your Turn 5

Alter the code to add a `prev_lifeExp` column that contains the life expectancy from the previous record.

(Hint: use "Transformation" cheatsheet, you want to offset elements by one)

Extra challenge: Why isn't this quite the 'life expectency five years ago'?

```{r}
gapminder %>%
  mutate()
```


***

# Take aways

* Extract cases with `filter()`  
* Filter using a list with `%in%`
* Make new variables, with `mutate()`  
* Connect operations with `%>%`  
