---
title: "Transform Data 1: manipulate cases"
output: html_notebook
---

<!-- This file by Charlotte Wickham (with some modifications by Martin Monkman) is licensed under a Creative Commons Attribution 4.0 International License, adapted from the orignal work at https://github.com/rstudio/master-the-tidyverse by RStudio. -->

```{r setup}
library(tidyverse)
library(gapminder)

```

This exercise introduces functions within the {dplyr} package. Note that the functions are (almost always) verbs, describing what the function does.

## Functions

|function |action |
| :--- | ---: |
| `filter()`    | *keep rows*|
| `mutate()`    | *create a new variable*|
| `group_by()`  | *declare subsets in data*|
| `select()`    | *keep variables (or drop them -var)*|
| `rename()`    | *renaming variables*|
| `summarize()` <br> or `summarise()` | *summarize the data, by groups if they have been declared*|
| `tally()`     | *counting (by groups if group_by() applied)*|
| `distinct()`  | *returns only rows that are unique*| 
| `case_when()` | *is used for "recoding" variable, often used with mutate()*|


## Arithmetic operators

| Operator | Description |
| : ----- | ---: |
| + | *plus* |
| - | *minus* |
| \* | *multiplication*  |
| / | *division* |
| ^ | *exponential* |


## Boolean operators

Boolean operators are a special type of operator that return TRUE or FALSE.  

| Operator | Description |
| : ----- | ---: |
| ==    | *equal, tests equality* |
| !=    | *not equal, tests inequality* |
| \>    | *greater than, tests greater than* (also >=) |
| <     | *less than, tests less than* (also <=) |
| %in%  | *contains, tests inclusion* |



---


Boolean operators can be combined with *and* or *or*.  

| Operator | Description |
| : ----- | ---: |
| `&`     | *and, returns true if preceeding and following are both true, else false* |
| &#124;  | *or, returns true if either preceeding and following are true, else false* |




## Useful functions

Some basic statistical functions  

| Operator | Description |
| : ----- | ---: |
| `min()` <br> `max()` | *minimum and maximum values* |
| `mean()` <br> `median()` <br> `mode()` | *measures of "average"* |
| `summary()`  | *a list of measures of central tendency* |
| `sum()`  | *adds up the values* |
| `count()`  | *number of cases* |


???

Source: Gina Reynolds,[Tidyverse in Action](https://github.com/EvaMaeRey/tidyverse_in_action)

Also see [R for Data Science, "Comparisions""](https://r4ds.had.co.nz/transform.html#comparisons) and ["Logical Operators"](https://r4ds.had.co.nz/transform.html#logical-operators)


## gapminder

For this exercise, we will use data from the {gapminder} package. For each of 142 countries, the package provides values for life expectancy, GDP per capita, and population, every five years, from 1952 to 2007.

It's a slice of data found at [gapminder.org](gapminder.org), which was founded by Hans Rosling. If you haven't already seen his videos, I suggest:

* [200 Years That Changed The World](https://www.gapminder.org/videos/200-years-that-changed-the-world/)

* [TED talk "The Best Stats You've Ever Seen"](https://www.ted.com/talks/hans_rosling_the_best_stats_you_ve_ever_seen)

and his book, [_Factfulness: Ten Reasons Why We're Wrong About the World--and Why Things Are Better Than You Think_](https://www.gapminder.org/factfulness-book/).


First, we load the data. Since the package is already installed into our R environment, we can just do the following:

```{r}
gapminder
```

## Your Turn 1: filter() rows

See if you can use the logical operators to manipulate our code below to show:

The data for Canada

```{r}
filter(gapminder, country == "New Zealand")
```

All data for countries in Oceania

```{r}
filter(gapminder, country == "New Zealand")
```

Rows where the life expectancy is greater than 82
```{r}
filter(gapminder, country == "New Zealand")
```


## Your Turn 2: multiple criteria

Use Boolean operators to alter the code below to return only the rows that contain:

* United States before 1970

```{r}

gapminder %>%
  filter(country == "New Zealand", 
         year > 2000)

```

*  Countries where life expectancy in 2007 is below 50 or over 75

```{r}


```



* Create a list object and return records for any of "New Zealand", "Canada" or "United States"


```{r}

```

* Return records for all continents except Europe


```{r}

```

## Your Turn 3: sorting

Find the records with the smallest population.
```{r}

```

Find the records with the largest GDP per capita.
```{r}

```


## Your Turn 4: creating new variables

Calculate a new variable with total GDP

```{r, error = TRUE}



```

Create a summary table with the population and GDP by continent for the year 2007  

```{r}



```

## Your Turn 5

Alter the code to add a `prev_lifeExp` column that contains the life expectancy from the previous record.

(Hint: use "Transformation" cheatsheet, you want to offset elements by one)

Extra challenge: Why isn't this quite the 'life expectency five years ago'?

```{r}
gapminder %>%
  mutate()
```


***

# Take aways

* Extract cases with `filter()`  
* Filter using a list with `%in%`
* Make new variables, with `mutate()`  
* Connect operations with `%>%`  


---

# Tidy data

> “Happy families are all alike; every unhappy family is unhappy in its own way.”<br> –– Leo Tolstoy
>“Tidy datasets are all alike, but every messy dataset is messy in its own way.”<br> –– Hadley Wickham


## Principles of tidy data

### 1. Each variable must have its own column.

### 2. Each observation must have its own row.

### 3. Each value must have its own cell.




From [_R for Data Science_, "Tidy Data"](https://r4ds.had.co.nz/tidy-data.html)

* see also [Hadley Wickham, 2014, “Tidy Data”, _Journal of Statistical Software_ 59 (1): 1–23](https://doi.org/10.18637/jss.v059.i10)


---

# Pivot tables with {tidyr}

Often we find ourselves dealing with untidy data. Or it might be tidy,  but in a structure that isn't particularly human-readable.

For example, if we were to filter the gapminder data to show changes in life expectancy in the Americas, with 15 year intervals.

```{r}

gapminder_Americas <- gapminder %>% 
  filter(continent == "Americas",
         year %in% c(1952, 1977, 1992, 2007))

gapminder_Americas

```

It's still a table with 100 rows!

But we can select only the countries, the years, and the life expectancy variable, and then _pivot_ the table structure so that the country is the row, the years are the columns, and the life expectancy values are under the years.

We use the {tidyr} function `pivot_wider()` to achieve this.

First, we create a revised version of the summary table with just the three variables we want in our table:

```{r}

gapminder_Americas <- gapminder_Americas %>% 
  select(country, year, lifeExp)

gapminder_Americas

```

Then we pivot:

```{r}

gm_pivot <- gapminder_Americas %>% 
  pivot_wider(names_from = year, values_from = lifeExp)

gm_pivot

```


We can return it to the original shape with the function `pivot_longer()`

```{r}

gm_pivot %>% 
  pivot_longer(cols = -country, names_to = "year", values_to = "life_exp")

```


### another pivot example

Is this data tidy?


```{r echo=FALSE}
# Elections Canada October 21, 2019 Federal Election Results
# https://enr.elections.ca/National.aspx?lang=e
# sourced 2019-11-06

CDN_elec_2019 <- tibble(party = c("Liberal", "Conservative", "Bloc Québécois", "New Democratic", "Green", "Independent"),
                        seats = c(157, 121, 32, 24, 3, 1),
                        votes = c(5915950, 6155662, 1376135, 2849214, 1162361, 75836))
# hahaha! It's easier to make a tidy tibble and untidy it than make a messy one to start with

CDN_elec_2019 <- CDN_elec_2019 %>%
  pivot_longer(cols = -party, names_to = "category", values_to = "number")
CDN_elec_2019
```


It violiates Principles #1 & #2

* there are two variables (seats and votes) combined into one column, and

* there are two rows (seats and rows) for each observation (in this case, party)


_Sometimes untidy data can be too long!_


### your turn

To tidy the table, we need to `pivot_wider()`

```{r}

# solution
CDN_elec_2019 %>%
  pivot_wider(names_from = category, values_from = number)

```



-30-
