---
title: "housing need"
output: html_notebook
---


```{r setup}
# tidyverse
library(tidyverse)
library(glue)

# utilities
library(here)
library(janitor)
```


## Data


[Core housing need, 2016 Census](https://www12.statcan.gc.ca/census-recensement/2016/dp-pd/chn-biml/index-eng.cfm) -- Dictionary, Census of Population, 2016


[Dictionary, Census of Population, 2016](https://www12.statcan.gc.ca/census-recensement/2016/ref/dict/households-menage037-eng.cfm)


[Data: Census tracts (CT) for CMAs and tracted CAs (1,516 KB)](https://www12.statcan.gc.ca/census-recensement/2016/dp-pd/chn-biml/CT_SR.csv) [CSV file]


CT

```{r read_data}
df_ct_housingneed <- read_csv(here("data", "CT_SR.csv"), 
                              n_max = 5641)

df_ct_housingneed <- clean_names(df_ct_housingneed)

```


Note: use of 

* `case_when()` : [A general vectorised if](https://dplyr.tidyverse.org/reference/case_when.html)

* `str_length()`: [The length of a string](https://stringr.tidyverse.org/reference/str_length.html)

```{r rename_vars}

df_ct_housingneed <- df_ct_housingneed %>% 
  rename(geo_code = "geographic_code_code_g_e9_ographique") %>% 
  mutate(geo_code = as.character(geo_code)) 

```





```{r}

df_ct_housingneed <- df_ct_housingneed %>% 
  mutate(geo_code_2 = case_when(
    str_length(geo_code) == 5 ~ paste(geo_code, ".00", sep = ""),
    str_length(geo_code) == 7 ~ paste(geo_code, ".00", sep = ""),
    TRUE ~ as.character(geo_code)
  ))


df_ct_housingneed <- df_ct_housingneed %>% 
  rename(rate_of_core_housing_need_2016 = rate_of_core_housing_need_2016_1_taux_des_besoins_imp_e9_rieux_en_mati_e8_re_de_logement_2016_1)
  



```


Now we have a list of all the Census Tracts in Canada.


## Join

To join it to the Prince George income by census tract table we made earlier, `df_ct_970`.

The first step is to filter to just the total category from the income data frame, then to join using the two different geo codes. 

```{r}

df_ct_970_housingneed <- df_ct_970 %>% 
  filter(household_type_cat == 1) %>% 
  left_join(df_ct_housingneed, by = c("geo_code_por" = "geo_code_2"))

df_ct_970_housingneed

```



## Modelling 

Statistical analysis is built into R's frame and engine, which means that things like correlation and regression analysis are straight-forward.

* see the [_IV Model_ section of _R for Data Science_](https://r4ds.had.co.nz/model-intro.html), particularly [Model Basics](https://r4ds.had.co.nz/model-basics.html)

In our example, we can test the relationship between household income and core housing need...our hypothesis is that as income increases, core housing need will fall.

```{r}

mod_housingneed <- lm(rate_of_core_housing_need_2016 ~ inc_median_total, data = df_ct_970_housingneed)


```


We can examine our model using `coef()` and (my go-to) `summary()`

```{r}

summary(mod_housingneed)

```


### plot the model

R gives us a quick way to visualize the relationship between two variables. With {ggplot2}, we can create a scatterplot using the `geom_point()` geography.


```{r}

ggplot(df_ct_970_housingneed, aes(x = inc_median_total, y = rate_of_core_housing_need_2016)) +
  geom_point()

```



```{r}

ggplot(df_ct_970_housingneed, aes(x = inc_median_total, y = rate_of_core_housing_need_2016)) +
  geom_point() +
  geom_smooth(method = lm)

```



The default in {ggplot2} is a solid line with the error range shown. But we can change those:


```{r}

ggplot(df_ct_970_housingneed, aes(x = inc_median_total, y = rate_of_core_housing_need_2016)) +
  geom_point() +
  geom_smooth(method = lm, se=FALSE, linetype="dashed", color="darkred")

```
























